---
title: "05_analysis_1"
format: 
  html:
    embed-resources: true
editor: visual
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
source("99_proj_func.R")
```

# Data analysis 1

In this part of the data analysis, we will be exploring the SIDER and Drugbank datasets.

Load merged data from Drugbank and SIDER.

```{r}
results_path = "../results/"
df <- read_tsv("../data/merged_data.tsv")

side_effect_rates <- df |>
  mutate(percentages = str_extract_all(freq_text, "\\d+\\.?\\d*(?=%)")) |>
  unnest_longer(percentages) |>
  mutate(percentages = as.numeric(percentages)) |>
  group_by(side_effect_PT) |>
  summarise(
    avg_rate = mean(percentages, na.rm = TRUE),
    median_rate = median(percentages, na.rm = TRUE),
    n_reports = n_distinct(`drugbank_id`),  # or however you define a "report"
    n_percentage_mentions = sum(!is.na(percentages)),
    .groups = "drop"
  ) |>
  filter(n_percentage_mentions > 0)
```

Let's just check what we're dealing with here

```{r}
side_effect_rates |> arrange(desc(n_reports))
```

Nausea, headache, vomiting, dizziness are all among the top reported side effects which seems to make sense.

```{r}
# Filter once so both correlation and plot use the exact same data
df_se <- side_effect_rates

# Compute and print the correlation
correlation <- cor(df_se$n_reports, df_se$avg_rate, use = "complete.obs")
print(paste("Correlation between number of reports and average frequency rate:", round(correlation, 3)))

# Plot
plot_correlation <- df_se |>
  ggplot(aes(x = n_reports, y = avg_rate)) +
  #ggplot(aes(x = n_percentage_mentions, y = avg_rate)) +
  geom_point(alpha = 0.5, color = "darkcyan") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Correlation: Number of Reports vs. Frequency Rate",
    subtitle = paste("Correlation coefficient:", round(correlation, 3)),
    x = "Number of Reports (Drug-Side Effect Pairs)",
    y = "Average Frequency Rate (%)"
  ) +
  theme_minimal()
ggsave(paste0(results_path,"Analysis1_reports_vs_freq.png"), plot = last_plot(), 
       width = 8, height = 6, dpi = 300)
print(plot_correlation)
```

Here we checked the average reported frequency of a side effect dependent on how many times it was reported. There may seem to be a small trend that very common side effects are more often reported than more niche side effects, and the variance in low number of reports is quite high. Common side effects might be easier to report since more data points will occur and it's more variable with few reports on rare data points.

```{r}
# Can we show correlation between illness treatment and side effect
# We filter for the top 20 items to keep the plot readable
top_indications <- df |>
  separate_rows(indication_PT, sep = ";\\s*") |>
  count(indication_PT, sort = TRUE) |>
  head(20) |>
  pull(indication_PT)

top_side_effects <- df |>
  count(side_effect_PT, sort = TRUE) |>
  head(20) |>
  pull(side_effect_PT)

plot_matrix <- df |>
  separate_rows(indication_PT, sep = ";\\s*") |>
  filter(
    indication_PT %in% top_indications,
    side_effect_PT %in% top_side_effects
  ) |>
  count(indication_PT, side_effect_PT) |>
  ggplot(aes(x = side_effect_PT, y = indication_PT, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Co-occurences") +
  labs(
    title = "Co-occurrences: Top Indications vs. Side Effects",
    x = "Side Effect",
    y = "Indication"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(paste0(results_path,"Analysis1_co_occ_indic.png"), plot = last_plot(), 
       width = 8, height = 6, dpi = 300)
print(plot_matrix)
```

This co-occurrence matrix is meant to show if specific common side effects are tied to specific treatment indications. The data is hard to analyze as we can see that Nausea and Vomiting is tied to Infection which also seems to be a common effect of Infections in general. This illustration seems to just show the most common reports. To combat this we can use the frequencies instead, which is in the next plot.

```{r}
plot_matrix_normalized_data <- df |>
  # extract % values
  mutate(percentages = str_extract_all(freq_text, "\\d+\\.?\\d*(?=%)")) |>
  # unnest so each percentage gets its own row 
  unnest_longer(percentages) |>
  mutate(percentages = as.numeric(percentages)) |>
  # remove NAs
  filter(!is.na(percentages)) |>
  # separate multiple indications
  separate_rows(indication_PT, sep = ";\\s*") |>
  # Filter to only top indications and side effects
  filter(
    indication_PT %in% top_indications,
    side_effect_PT %in% top_side_effects
  ) |>
  
  # group by indication–side effect pair
  group_by(indication_PT, side_effect_PT) |>
  
  # get the stats
  summarise(
    avg_rate = mean(percentages, na.rm = TRUE),
    n_percentage_mentions = n(),
    n_drugs_contributing = n_distinct(drugbank_id),
    .groups = "drop"
  )
```

```{r}
plot_matrix_normalized <- plot_matrix_normalized_data |>
  ggplot(aes(x = side_effect_PT, y = indication_PT, fill = avg_rate)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Avg. Frequency(%)") +
  labs(
    title = "Co-frequencies: Top Indications vs. Side Effects",
    x = "Side Effect",
    y = "Indication"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(paste0(results_path,"Analysis1_co_occ_indic_freq.png"), plot = last_plot(), 
       width = 8, height = 6, dpi = 300)
# show plot
plot_matrix_normalized
```

This frequency matrix shows instead where a large % of side effects are experienced. In drugs that are used to treat breast cancer we often see Diarrhoea, Nausea, Feeling abnormal and Vomiting which may be expected which such a serious illness. Neoplasm is tumor growth and we see frequent Nausea and Gastrointenstinal pain, which again makes sense. We can also see that Hypertension rarely leads to any of these side effects when treated. Some of the serious diseases seem to be more yellow on the chart which might make sense as a more side effects may be allowed.

```{r}
# Get top targets and side effects
top_targets <- df |>
  mutate(uniprot_first = str_extract(uniprot_names, "^[^;]+")) |> 
  count(uniprot_first, sort = TRUE) |>
  head(20) |>
  pull(uniprot_first)

top_side_effects <- df |>
  count(side_effect_PT, sort = TRUE) |>
  head(20) |>
  pull(side_effect_PT)

# Plot heatmap
plot_targets_se <- df |>
  mutate(uniprot_first = str_extract(uniprot_names, "^[^;]+")) |>
  filter(
    uniprot_first %in% top_targets,
    side_effect_PT %in% top_side_effects
  ) |>
  count(uniprot_first, side_effect_PT) |>
  mutate(uniprot_first = str_trunc(uniprot_first, 30)) |>
  ggplot(aes(x = side_effect_PT, y = uniprot_first, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Co-occurrences") +
  labs(
    title = "Co-occurrences: Protein Targets vs. Side Effects",
    x = "Side Effect",
    y = "Protein Target"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(paste0(results_path,"Analysis1_co_occ_targets.png"), plot = last_plot(), 
       width = 8, height = 6, dpi = 300)
print(plot_targets_se)
```

Here we can see some co-occurrences between what biological target/receptor is treated and what type of side effect is experienced, but this requires a further deep dive.

```{r}
# plot biological targets vs side effects
# Calculate per-drug metrics
drug_targets_se <- df |>
  group_by(drugbank_id, common_name) |>
  summarise(
    # Count unique targets
    n_targets = n_distinct(unlist(str_split(uniprot_names, ";\\s*"))),
    # Count unique side effects
    n_side_effects = n_distinct(side_effect_PT),
    .groups = "drop"
  ) |>
  filter(!is.na(n_targets), n_targets > 0)

# Compute correlation
cor_targets <- cor(drug_targets_se$n_targets, drug_targets_se$n_side_effects, 
                   use = "complete.obs")

# 
plot_targets_vs_se <- drug_targets_se |>
  ggplot(aes(x = n_targets, y = n_side_effects)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Number of Biological Targets vs. Number of Side Effects",
    subtitle = paste("Correlation:", round(cor_targets, 3)),
    x = "Number of Protein Targets",
    y = "Number of Unique Side Effects"
  ) +
  theme_minimal()

ggsave(paste0(results_path,"Analysis1_targets_se_correl.png"), plot = last_plot(), 
       width = 8, height = 6, dpi = 300)
print(plot_targets_vs_se)
```

We see no explicit relation between the amount of protein targets a given drug has versus the amount of side effects it gives, which is surprising.

```{r}
# plot indications vs side effects
drug_indications_se <- df |>
  group_by(drugbank_id, common_name) |>
  summarise(
    # Count unique indications
    n_indications = n_distinct(unlist(str_split(indication_PT, ";\\s*"))),
    # Count unique side effects
    n_side_effects = n_distinct(side_effect_PT),
    .groups = "drop"
  ) |>
  filter(!is.na(n_indications), n_indications > 0)

# Compute correlation
cor_indications <- cor(drug_indications_se$n_indications, drug_indications_se$n_side_effects,
                       use = "complete.obs")

# Plot
plot_indications_vs_se <- drug_indications_se |>
  ggplot(aes(x = n_indications, y = n_side_effects)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Indications vs. Side Effects",
    subtitle = paste("Correlation:", round(cor_indications, 3)),
    x = "Number of Indications",
    y = "Number of Unique Side Effects"
  ) +
  theme_minimal()

ggsave(paste0(results_path,"Analysis1_indications_se_correl.png"), plot = last_plot(), 
       width = 8, height = 6, dpi = 300)
print(plot_indications_vs_se)
```

We do see some amount of correlated behavior between how many indications a drug targets or treats and how many side effects are listed. This is surprising compared to the former plot, but might be explained by the fact that more indications on a drug potentially means that it might have been studied more and therefore more side effects appeared.
